---
title: "NYPD Shootings EDA"
author: "Swapnil Bhatta"
date: "2022-11-15"
output:
  pdf_document: default
  html_document: default
---
```{r Install of Packages, include = FALSE}

#!!!READ  BEFORE RUNNING: Comment out below install of packages if you don't want them installed through this method. Only completed to ensure the document would knit first try.

#if(!require(tidyverse)) install.packages("tidyverse",repos = "http://cran.us.r-project.org")

#if(!require(lubridate)) install.packages("lubridate",repos = "http://cran.us.r-project.org")

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## How to get the data

Navigate to the dataset link https://catalog.data.gov/dataset and search for a dataset titled NYPD Shooting Incident Data (Historic). The repo has the csv file, but the script  automatically downloadeds it from the website when ran.

```{r read data, message = FALSE}
nypd_url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
police_shootings <- read.csv(nypd_url,
                             header=TRUE,
                             sep=",")
```

```{r Summary}
summary(police_shootings)
```
## What cleaning process was used? 
- Changed the type for OCCUR_DATE to date
- Removed unwanted columns


```{r Data select}
# Remove unwanted columns
police_shootings <- select(police_shootings, -c(LOCATION_DESC, STATISTICAL_MURDER_FLAG, PERP_AGE_GROUP, X_COORD_CD, Y_COORD_CD, Lon_Lat))
```


```{r Date add column}
# Convert string to datetime
police_shootings$ OCCUR_DATE <- mdy(police_shootings$ OCCUR_DATE)
```

## Daily frequency of shootings


```{r Data further Preprocess}
# Add columns for monthly, yearly averages
police_shootings <- police_shootings %>% 
    mutate(YEAR_MONTH = floor_date(police_shootings$ OCCUR_DATE, 'month')) %>% 
    mutate(MONTH = strftime(police_shootings$ OCCUR_DATE, format='%m'))
```


```{r Incidents plot}
date_value_counts <- police_shootings %>% count(YEAR_MONTH)
ggplot(date_value_counts, aes(x=YEAR_MONTH, y=n)) +
                        geom_line(color = "cornflowerblue", size=1) +
                        labs(x = "Date", y = "Number of Incidents", title='Shooting Incidents - Time Series')
```

```{r Incidents narrow}
# Checking if some incidents are reported additional times for multiple shootings
incident_counts <- police_shootings %>% count(INCIDENT_KEY) 
incident_counts[order(incident_counts$ n, decreasing = TRUE),] %>% head(10)
```

We see that an incident can have multiple data points, plotting only unique counts confirms that there might've been some outliers 


```{r Incidents narrow Plot}
distinct_year_month <- police_shootings %>% group_by(YEAR_MONTH) %>%
                    summarize(distinct_incident = n_distinct(INCIDENT_KEY))
ggplot(distinct_year_month, aes(x=YEAR_MONTH, y=distinct_incident)) + 
                geom_line(color = "darkorange1") +
                labs(x = "Date", y = "Number of Distinct Incidents", title='Distinct Shooting Incidents - Time Series')  
```

Visualizing overlap of incidents


```{r Overlap visualization}
ggplot(date_value_counts, aes(x=YEAR_MONTH, y=n)) + 
                        geom_line(color = "cornflowerblue") + 
                        geom_line(data = distinct_year_month, aes(x=YEAR_MONTH, y=distinct_incident), color = "darkorange1") +
                    labs(x = "Date", y = "Number of Incidents", title='Shooting Incident Overlaps - Time Series') 
```


Hourly breakdown for shooting incidents


```{r Distinct hours selection}
distinct_hours <- police_shootings %>% group_by(OCCUR_TIME) %>%
                    summarize(distinct_incident = n_distinct(INCIDENT_KEY))
ggplot(distinct_hours, aes(x=OCCUR_TIME,
                              y=distinct_incident)) + geom_line(color = "blue")+
                    labs(x = "Date", y = "Number of Incidents", title='Shooting Incident Overlaps - Time Series')+
                    labs(x = "Time (Hour)", y = "Number of Incidents", title='Shooting Incident - Hourly Time Series') 
```


Most shooting incidents take place after hours, between 8 PM and 4 AM

```{r distinct hours plot}
distinct_lat_long <- police_shootings %>% group_by(Latitude, Longitude, BORO) %>%
                    summarize(distinct_incident = n_distinct(INCIDENT_KEY))

ggplot(distinct_lat_long, aes(x=Latitude, y=Longitude, color=BORO)) + 
        geom_point(aes(size=distinct_incident)) +
        geom_point(shape = 1, aes(size=distinct_incident), alpha = 0.1, colour = "black") +
                    labs(x = "Latitude", y = "Longitude", title='Scatter Plot Showing NYC Boroughs, Incidents') 
```


We can visualize the approximate location of boroughs and the size of incidents for various locations


```{r gender data selection}
distinct_gender <- police_shootings %>% group_by(VIC_SEX) %>%
                    summarize(distinct_incident = n_distinct(INCIDENT_KEY))
ggplot(distinct_gender, aes(x="", y=distinct_incident, fill=VIC_SEX, label = distinct_incident)) + 
                geom_bar(stat="identity", width=1, color="white") +
                          coord_polar("y", start=0) + theme_void() +
                geom_text(aes(label = distinct_incident), position = position_stack(vjust = 0.5), size=5) +
                    labs(title='Gender Breakdown for Incidents') 
```


## Add a linear model to see if number of shootings have been increasing


```{r linear model}
linear_incidents = lm(distinct_incident~YEAR_MONTH, data=distinct_year_month)
```


```{r linear model plot}
ggplot(distinct_year_month, aes(x=YEAR_MONTH,
                                y=distinct_incident, group = 1)) + 
                                geom_line(color = "darkorange1") +
                                geom_smooth(method='lm', se = FALSE) +
                    labs(x = "Date", y = "Number of Distinct Incidents", title='Distinct Shooting Incidents with Linear Fit') 
```

The linear model suggests the number of shooting has decreased


```{r linear summary}
summary(linear_incidents)
```


The R value suggests a poor fit, which is expected given the seasonality in the data and the outlier for the pandemic years.


```{r monthly select}
distinct_monthly <- police_shootings %>% group_by(MONTH) %>%
                                summarize(distinct_incident = n_distinct(INCIDENT_KEY))
ggplot(distinct_monthly, aes(x=MONTH,
                                y=distinct_incident, group = 1)) + 
                                geom_line(color = "darkorange1") +
                    labs(x = "Month", y = "Number of Distinct Incidents", title='Distinct Shooting Incidents Aggregated Monthly') 
```


Summer months have the highest number of shootings.

## Possible Bias

The collected data itself can be biased due to higher police presence and incidents in communities that have predominantly minority population, further exploration of the PERP_RACE fields and understanding of New York City's racial distribution areas can help shed more light into it.

